Recursive Solutions to First-Order Model Counting

### 2022-07-30 doc/paper/paper.tex
Transfer changes from the thesis

\captionsetup[subfigure]{width=\linewidth}
\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{identifyRecursion}
\SetKwFunction{generateMaps}{generateMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. WFOMC emerged as the dominant approach
to \emph{lifted (probabilistic) inference}. Lifted inference techniques exploit
symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in theory \citep{DBLP:journals/jair/Kuzelka21}.}
We claim that this shortcoming is due to the inability of these algorithms to
construct recursive solutions. The topic of recursion in the context of WFOMC
has been studied before but in limited ways.
\Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data which
is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

% \begin{itemize}
%   \item But their code only conjectures stuff on data. My conjectures are
%         guaranteed to always be true.
%   \item everything has to be expressible in \Ctwo{}, i.e., the two-variable
%         fragment of FOL with counting quantifiers on one domain.
%   \item They only consider P-recursive functions, which means that, e.g.,
%         $f(n) = f(n-1)f(n-2)$ would not be allowed.
% \end{itemize}

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances in a lifted
manner. The ideas presented in this paper are implemented in
\textsc{Crane}---an extension of the arguably most well-known WFOMC algorithm
\textsc{ForcLift} written in Scala. \textsc{ForcLift} works in two stages:
compilation and evaluation/propagation.\footnote{There is also an intermediate
  stage called \emph{smoothing} that takes place between compilation and
  evaluation. As our changes to smoothing are quite elementary, we do not
  discuss them to not distract the reader from the main contributions of this
  paper.} In the first part, various \emph{(compilation) rules} are applied to
the input (or some derivative) formula, gradually constructing a circuit. In the
second part, the weights of the instance (and sometimes constants) are
propagated through the circuit, computing the WMC\@.

@@

TODO: improve the wording of the paragraph below.

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits. The cycles in
these graphs represent recursive calls. Suppose the original formula $\phi$
depends on a domain of size $n \in \mathbb{N}$. Our generalised version of
domain recursion transforms $\phi$ into a different formula $\psi$ that depends
on a domain of size $n-1$. After some number of subsequent transformations, the
algorithm identifies that a solution to $\psi$ can be constructed in part by
finding a solution to a version of $\phi$ where the domain of size $n$ is
replaced by a domain of size $n-1$. Recognising $\phi$ from before, we can add a
cycle-forming edge to the graph, which can be interpreted as function $f$
relying on $f(n-1)$ to compute $f(n)$.

@@

To construct such graphs, we introduce a novel search algorithm that replaces
greedy search used by \textsc{ForcLift} and three compilation rules that work
together to construct cyclic graphs. We begin with \cref{sec:recprelims} where
we define the representation used for sentences in FOL as well as discuss
caching and some notational conventions. Then, at the beginning of
\cref{sec:methods} we formally define the graphs that replace circuits in
representing a solution to a (W)FOMC problem and discuss some related notions.
\Cref{sec:rules} introduces the new compilation rules:
\begin{itemize}
  \item a generalisation of domain recursion by \citet{DBLP:conf/nips/Broeck11},
  \item \emph{constraint removal}---a formula transformation rule that removes
        the constraints introduced by generalised domain recursion (GDR),
        introducing a new (smaller) domain instead,
  \item and a rule for creating recursive calls.
\end{itemize}
In a typical solution produced by \textsc{Crane}, these rules are executed in
exactly this order (interspersed with some other rules), with the edge that
represents a recursive call pointing back to the node introduced by GDR\@. In
\cref{sec:interpret}, we informally discuss how a graph can be interpreted as a
collection of (potentially recursive) functions. Finally, in \cref{sec:results}
we compare \textsc{ForcLift} and \textsc{Crane} on a range of function-counting
problems. We show that \textsc{Crane} performs as well as \textsc{ForcLift} on
the instances that were already solvable by \textsc{ForcLift} but is also able
to handle most instances that \textsc{ForcLift} fails on as well.

% \Cref{sec:search} then describes our new search algorithm that makes fewer
% assumptions than \textsc{ForcLift} about the right order in which compilation
% rules should be applied.

% \begin{itemize}
%   \item What we do with the compilation result is different (see below)
%   \item evaluation/interpretation: algebraic simplification, dynamic programming
%   \item limitation: function extraction and simplification need to be automated,
%         same for finding base cases
% \end{itemize}

% Leftovers
% \begin{itemize}
%   \item compilation works by always having a formula as input and
%         adding nodes to the circuit/FCG that constitute the solution the that
%         formula.
%   \item Workflow after
%   \begin{itemize}
%     \item Compile the formula to a graph
%     \item Extract the definitions of functions
%     \item Simplify
%     \item Supplement with base cases
%     \item Evaluate to get the answer
%   \end{itemize}
% \end{itemize}

@@

Our representation of FOMC instances is heavily based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} is able to translate
sentences in a variant of function-free many-sorted FOL with equality to this
internal format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$
and terms $t_{1}, \dots, t_{n}$. Here, $n \in \mathbb{N}_0$ is the \emph{arity}
of $p$. A \emph{term} is either a constant or a variable. A \emph{literal} is
either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains; note
that this set expands during compilation.

@@

Similarly to variables in \cref{def:clause}, all constants are (implicitly)
mapped to domains, and each $n$-ary predicate is implicitly mapped to a sequence
of $n$ domains. For constant $x$, predicate $p$, and domains $a$ and $b$, we
write, e.g., $x \in a$ and $p \in a \times b$ to denote that $x$ is associated
with $a$, and $p$ is associated with $a$ and $b$ (in that order).

@@

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto a, Y \mapsto b, Z \mapsto a \,\})
          \end{multlined}
  \end{align*}
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $a$ and $b$. Then
  $\Vars(\{\, (Y, Z) \,\}) = \{\, Y, Z \,\}$, and
  $\Vars(c_{1}) = \Vars(c_{2}) = \{\, X, Y, Z \,\}$. Based on the domain maps of
  $c_{1}$ and $c_{2}$, we can infer that $p \in a \times b$. All variables (in
  both clauses) that occur as the first argument to $p$ are in $a$, and likewise
  all variables that occur as the second argument to $p$ are in $b$. Therefore,
  $\phi$ `type checks' as a valid formula.
\end{example}

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by \citet{DBLP:conf/ijcai/BroeckTMDR11}. First, we decouple
variable-to-domain assignments from constraints and move them to a separate
function $\delta_{c}$ in \cref{def:clause}. Formalising these assignments as a
function unveils the (previously implicit) assumption that each variable must be
assigned to a domain. Second, while \citet{DBLP:conf/ijcai/BroeckTMDR11} allow
for equality constraints and constraints of the form $X \not\in d$ for some
variable $X$ and domain $d$, we exclude such constraints simply because they are
not needed.

One can read a formula as in \cref{def:formula} as a sentence in a FOL\@.
All variables in a
clause are implicitly universally quantified (but note that variables are never
shared among clauses), and all clauses in a formula are implicitly linked by
conjunction. Thus, formula $\phi$ from \cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in a\text{. }\forall Y, Z \in b\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in a\text{. }\forall Y \in b\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|a| = |b| = 2$. There are
  $2^{2 \times 2} = 16$ possible relations between $a$ and $b$. Let us count how
  many of them satisfy the conditions imposed on predicate $p$. The empty
  relation does. All four relations of cardinality one do too. Finally, there
  are two relations of cardinality two that satisfy the conditions as well.
  Thus, the FOMC of $\phi$ (when $|a| = |b| = 2$) is 7. Incidentally, the FOMC
  of $\phi$ counts partial injections. We will continue to use the problem of
  counting partial injections (and the formula from \cref{example:first}
  specifically) as the main running example throughout the paper.
\end{example}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{a \gets a \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{b^\top \subseteq b}$}
      child {node {$\CR_{a \mapsto a'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          % _{(\emptyset, \{\, (X, Y) \,\}, \{\, X \mapsto b^\top, Y \mapsto b^\top \,\})}
          child {node (ref) {$\Reff_{\{\, a \mapsto a', b \mapsto b^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{b^\top \subseteq b}$ denotes set-disjunction, $\land$ denotes
    conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit nodes whose only arithmetic effect is multiplication by
    one. Some of these nodes play an important role in the weighted version of
    the problem whereas others are remnants of the interaction between
    compilation rules and the way in which \textsc{ForcLift} handles existential
    quantifiers.}\label{fig:examplefcg}
\end{figure}

>>>>

\captionsetup[subfigure]{width=\linewidth}
\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{identifyRecursion}
\SetKwFunction{generateMaps}{generateMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \citep{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as a WMC instance---in fact, the WFOMC algorithm
by \citet{DBLP:conf/ijcai/BroeckTMDR11} can do so when it is unable to solve the
instance in a lifted manner. WFOMC emerged as the dominant approach to
\emph{lifted (probabilistic) inference}. Lifted inference techniques exploit
symmetries in probabilistic models by reasoning about sets rather than
individuals \citep{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \citep{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\citep{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\citep{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \citep{DBLP:journals/ml/KimmigMG15} and statistical
relational \citep{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\citep{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. \Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data
which is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift} written in Scala. \textsc{ForcLift}
works in two stages: compilation and evaluation/propagation.\footnote{There is
  also an intermediate stage called \emph{smoothing} that takes place between
  compilation and evaluation. As our changes to smoothing are quite elementary,
  we do not discuss them to not distract the reader from the main contributions
  of this paper.} In the first part, various \emph{(compilation) rules} are
applied to the input (or some derivative) formula, gradually constructing a
circuit. In the second part, the weights of the instance (and sometimes
constants) are propagated through the circuit, computing the WMC\@.

@@

TODO: remove the references to the figure.

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits. The cycles in
these graphs represent recursive calls. See \cref{fig:after} for an illustration
of an example scenario. Suppose the original formula $\phi$ depends on a domain
of size $n \in \mathbb{N}$. \emph{Generalised domain recursion} (GDR), one of
the new compilation rules, transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another new compilation rule recognizes that the
resulting formula $\phi(n-1)$ is just like the input formula $\phi(n)$ except it
refers to a different domain. This observation allows us to add a cycle-forming
edge to the graph (highlighted in blue in \cref{fig:after}), which can be
interpreted as function $f$ relying on $f(n-1)$ to compute $f(n)$. To construct
such graphs, we also introduce a novel search algorithm that replaces the greedy
search used by \textsc{ForcLift}.

@@

\Cref{sec:recprelims} defines the representation used for sentences in FOL as
well as discuss caching and some notational conventions. At the beginning of
\cref{sec:methods}, we formally define the graphs that replace circuits in
representing a solution to a (W)FOMC problem and discuss some related notions.
\Cref{sec:rules} introduces the new compilation rules. \Cref{sec:search} then
describes our new search algorithm that makes fewer assumptions than
\textsc{ForcLift} about the right order in which compilation rules should be
applied. In \cref{sec:interpret}, we discuss how a graph can be interpreted as a
collection of (potentially recursive) functions. Finally, in \cref{sec:results}
we compare \textsc{ForcLift} and \textsc{Crane} on a range of function-counting
problems. We show that \textsc{Crane} performs as well as \textsc{ForcLift} on
the instances that were already solvable by \textsc{ForcLift} but is also able
to handle most of the instances that \textsc{ForcLift} fails on.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} is able to translate
sentences in a variant of function-free many-sorted FOL with equality to this
internal format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$
and terms $t_{1}, \dots, t_{n}$. Here, $n \in \mathbb{N}_0$ is the \emph{arity}
of $p$. A \emph{term} is either a constant or a variable. A \emph{literal} is
either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains.
Initially, $\mathcal{D}$ contains all domains mentioned by the input (W)FOMC
instance. During compilation, new domains added to $\mathcal{D}$ by some of the
compilation rules.

@@

Similarly to variables in \cref{def:clause}, all constants are (implicitly)
mapped to domains, and each $n$-ary predicate is implicitly mapped to a sequence
of $n$ domains. For constant or variable $x$, predicate $p$, and domains
$\Gamma$ and $\Delta$, we write, e.g., $x \in \Gamma$ and
$p \in \Gamma \times \Delta$ to denote that $x$ is associated with $\Gamma$, and
$p$ is associated with $\Gamma$ and $\Delta$ (in that order).

@@

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $\Gamma$ and
  $\Delta$. Then $\Vars(\{\, (Y, Z) \,\}) = \{\, Y, Z \,\}$, and
  $\Vars(c_{1}) = \Vars(c_{2}) = \{\, X, Y, Z \,\}$. Based on the domain maps of
  $c_{1}$ and $c_{2}$, we can infer that $p \in \Gamma \times \Delta$. All
  variables (in both clauses) that occur as the first argument to $p$ are in
  $\Gamma$, and likewise all variables that occur as the second argument to $p$
  are in $\Delta$. Therefore, $\phi$ `type checks' as a valid formula.
\end{example}

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by \citet{DBLP:conf/ijcai/BroeckTMDR11}. First, we decouple
variable-to-domain assignments from constraints and move them to a separate
function $\delta_{c}$ in \cref{def:clause}. Formalising these assignments as a
function unveils the (previously implicit) assumption that each variable must be
assigned to a domain. Second, while \citet{DBLP:conf/ijcai/BroeckTMDR11} allow
for equality constraints and constraints of the form $X \not\in \Delta$ for some
variable $X$ and domain $\Delta$, we exclude such constraints simply because
they are not needed.

One can read a formula as in \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by conjunction. Thus, formula $\phi$ from \cref{example:first}
reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\begin{example}
  Let $\phi$ be as in \cref{example:first} and let $|\Gamma| = |\Delta| = 2$.
  There are $2^{2 \times 2} = 16$ possible relations between $\Gamma$ and
  $\Delta$. Let us count how many of them satisfy the conditions imposed on
  predicate $p$. The empty relation does. All four relations of cardinality one
  do too. Finally, there are two relations of cardinality two that satisfy the
  conditions as well. Thus, the FOMC of $\phi$ (when $|\Gamma| = |\Delta| = 2$)
  is 7. Incidentally, the FOMC of $\phi$ counts partial injections. We will
  continue to use the problem of counting partial injections (and the formula
  from \cref{example:first} specifically) as the main running example throughout
  the paper.
\end{example}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit nodes whose only arithmetic effect is multiplication by
    one. Some of these nodes play an important role in the weighted version of
    the problem whereas others are remnants of the interaction between
    compilation rules and the way in which \textsc{ForcLift} handles existential
    quantifiers.}\label{fig:examplefcg}
\end{figure}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Note that
  imposing an ordering on outgoing edges is just a limited version of edge
  labelling.} We denote an FCG as $G = (V, s, N^+, \tau)$, where $V$ is the set
of nodes, and $s \in V$ is the unique source. Function $N^+$ maps each node in
$V$ to a \emph{list} of its direct successors. Node labels consist of two parts:
the \emph{type} and the \emph{parameters}. To avoid clutter, we leave the
parameters implicit and let $\tau$ denote the node-labelling function that maps
each node in $V$ to its type. For each node $v \in V$, the length of list
$N^+(v)$ (i.e., the out-degree of $v$) is determined by its type $\tau(v)$. Most
of the types are as in previous work
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}. The type for
non-tree-like edges (denoted by $\Reff$), while used before, is extended to
contain the information necessary to support recursive calls. We also add three
new types:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
When drawing an FCG, we order outgoing edges from left to right, write node
labels directly on the nodes, and omit irrelevant labels and/or parameters. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1
(i.e., $|N^+(s)| = 1$), label
$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$ (i.e.,
$\tau(s) = \GDR$).
